---
title: "Reproducible Research: Peer Assessment 1"
author: "Irina Telina"
date: "16.07.2015"
output: 
    html_document:
        keep_md: true
---
```{r echo=FALSE, results="hide"}  
Sys.setlocale("LC_TIME","English")
require(knitr)
opts_chunk$set(echo = TRUE, fig.path = "figure-html/")
``` 
### Loading the data:

```{r loading, echo=TRUE}  
df <- read.csv("./activity.csv", header = TRUE, stringsAsFactors = FALSE)  
```

## What is mean total number of steps taken per day?

```{r part1, echo=TRUE, message=FALSE}  
library(dplyr)  
dfDay <- group_by(df, date)  
TotalStepsDay <- summarize(dfDay, total_steps = sum(steps, na.rm = TRUE))  
hist(TotalStepsDay$total_steps, main = "The total number of steps taken each day", xlab = "Steps")  
mean_steps <- round(mean(TotalStepsDay$total_steps), digits = 2)    
median_steps <- median(TotalStepsDay$total_steps)  
```

The mean of the total number of steps taken per day is **`r mean_steps`**.
The median of the total number of steps taken per day is **`r median_steps`**.

## What is the average daily activity pattern?

```{r part2, echo=TRUE}
dfInterval <- group_by(df, interval)
AvgStepsInterval <- summarise(dfInterval, avg_steps = mean(steps, na.rm = TRUE))
plot(AvgStepsInterval$interval, AvgStepsInterval$avg_steps, type = "l", main = "The average number of steps", xlab = "5-minute interval", ylab = "Averaged number of steps across all days")
max_steps <- max(AvgStepsInterval$avg_steps)  
interval_max_step <- round(AvgStepsInterval[AvgStepsInterval$avg_steps == max_steps,1], digits = 2)  
```

The 5-minute interval # **`r interval_max_step`** contains the maximum number of steps, on average across all the days in the dataset.

## Imputing missing values

The total number of missing values in the dataset is **`r sum(is.na(df$steps))`**.

Filling in the missing values with the means for 5-minute intervals:
```{r part3, echo=TRUE}
df1 <- df
for (i in seq_along(df1$steps)) 
{
    if(is.na(df1[i,"steps"])) 
    {
        df1[i,"steps"] <- AvgStepsInterval[AvgStepsInterval$interval == df1[i,"interval"], "avg_steps"]
    }
}

```

Making a histogram of the total number of steps taken each day:
```{r part4, echo=TRUE}
dfDay1 <- group_by(df1, date)  
TotalStepsDay1 <- summarize(dfDay1, total_steps = sum(steps))  
hist(TotalStepsDay1$total_steps, main = "The total number of steps taken each day", xlab = "Steps")  
mean_steps1 <- mean(TotalStepsDay1$total_steps) 
median_steps1 <- median(TotalStepsDay1$total_steps)   
```

The mean of the total number of steps taken per day is **`r sprintf("%.2f", mean_steps1)`**.  
The median of the total number of steps taken per day is **`r sprintf("%.2f", median_steps1)`**.  
These values differ from the estimates from the first part of the assignment:  
- the mean of the total number of steps taken per day was `r mean_steps`;  
- the median of the total number of steps taken per day was `r median_steps`.  

The impact of imputing missing data on the estimates of the total daily number of steps consists in smoothing of the total number of steps distribution. 

## Are there differences in activity patterns between weekdays and weekends?

```{r part5, echo=TRUE}
df1$date <- strptime(df1$date, "%Y-%m-%d")
df1 <- mutate(df1, dayF = factor((weekdays(df1$date, abbreviate = TRUE) %in% c("Sat", "Sun")), labels = c("weekday", "weekend")))
AvgStepsInterval1 <- df1 %>% select(-date) %>% group_by(dayF, interval) %>% summarise(avg_steps=mean(steps))
library("ggplot2")
g <- ggplot(AvgStepsInterval1, aes(interval, avg_steps))
g + geom_line() + 
    facet_grid(dayF~.) + 
    labs(title = "The average number of steps", x = "5-minute interval", y = "Averaged number of steps across weekdays or weekends")
```
